Gaussian Process Regression for astronomical time-series

### 2022-11-15 src/tex/ms.tex
[showyourwork] overleaf sync

In the previous section, we presented an overview of GP methods, and the key mathematical details. In this section, we will dive deeper into some of the practical decisions that arise when using GPs.
The two core elements of a GP model are the mean function $m(t;\,\meanparams)$, and the ``kernel'' or ``covariance'' function $k(t_i,\,t_j;\,\hyperparams)$. In contrast to the way GP models are typically presented in the machine learning literature, in astrophysics they will often---but not always---include non-trivial mean functions.
For example, in the case study discussed in \autoref{sec:transit}, the mean function $m(t;\,\meanparams)$ is a physical transit model that is a function of the orbital parameters of the system, and includes a realistic limb-darkening model \citep{2002ApJ...580L.171M}.
However, in this review we will not discuss the mean function in detail, focusing instead primarily on the kernel function, since that is unique to GP modelling.
All this being said, in our experience, new users of GP models will often focus and worry more than necessary about the choice of kernel function for their problem.
As with any probabilistic modelling problem, there are several well-defined workflows for motivating, selecting, and validating your choice of kernel function.
In this section we walk through this process in detail.

@@

The kernel function can be any positive scalar function that gives rise to a positive semi-definite covariance matrix over the input domain, but some will be more useful than others.
Given this large decision space, you may be wondering how to choose the right kernel function for your specific problem.
If you are very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel functions have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that would naturally motivate their use in certain circumstances.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can be recast as a GP model with a specific kernel function.
% \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

Even if you do not have a formal physics-based justification for your model, you may be able to identify the key scales of your problem and design a kernel function that captures these features.
The usual approach to this problem is to take sums and products of commonly used kernel functions to construct a set of models that have the desired covariance structure, and then combine or select between these choices.
For example, if the problem of interest is expected to involve only a single non-periodic timescale, you could list all the two-parameter non-periodic kernel functions and use a numerical model selection technique as described below.

@@

When selecting a kernel function, it can be useful to generate samples from this implied prior distribution over functions to get a qualitative sense of the properties of the kernel.
In practice, this is done by choosing a grid of times $\boldsymbol{t} = \{t_i\}$, evaluating the elements of the covariance matrix
\begin{equation}
  K_{ij} = k(t_i,\,t_j;\,\hyperparams) \quad,
\end{equation}
and then generating a multivariate Gaussian sample with this covariance.
As an example, \autoref{fig:samples} shows several prior samples for three different kernel functions from \autoref{tab:kernels}, with a range of length scales $\lengthscale$ and amplitudes $\amplitude$.
In this figure, we can see some qualitative differences between the kernel functions---namely that the samples become ``smoother'' from left to right---and we can see how the characteristic input and output scales of the allowed functions change with the hyperparameters.

@@

\begin{figure}[ht]
  \centering
  \script{kernel-ops.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/kernel-ops1.pdf}
  \includegraphics[width=0.7\linewidth]{figures/kernel-ops2.pdf}
  \caption{A demonstration of the effects of differentiation on a GP model.
  Each row shows the same figures a different base kernel: \emph{(top)} a squared exponential kernel ($\lengthscale = 1$), and \emph{(bottom)} the product of a squared exponential kernel ($\lengthscale_1 = 2.5$) with a sine squared exponential kernel ($\Gamma = 1$; $\lengthscale_2 = 3.5$).
  In each row, the leftmost panel shows the kernel function and its first and second time derivatives, which are used to evaluate the covariance matrix for a GP model for a process and its derivative.
  The right-hand panel shows a realization from the process ($f(t)$, top) and its time derivative ($\dot{f}(t)$, bottom) generated by jointly sampling from the GP including the covariance between $f(t)$ and $\dot{f}(t)$.}
  \label{fig:kernel-ops}
\end{figure}

@@

When the input coordinates have some geometric interpretation (e.g., spatial coordinates) simple isotropic distance metrics, like those presented so far, are generally well-justified.
However, when the input coordinates are heterogeneous with, for example, different units, it is often more appropriate to use a more general distance metric.
In particular, you may want to fit for a different length scale parameter $\lengthscale_d$ for each input dimension $d$:
\begin{equation}\label{eq:ard-distance}
  \rho(\boldsymbol{x}_i,\boldsymbol{x}_j) = \sqrt{\sum_{d=1}^D \left(\frac{x_{id} - x_{jd}}{\lengthscale_d}\right)^2}
\end{equation}
or even include support for covariance between different input dimensions:
\begin{equation}
  \rho(\boldsymbol{x}_i,\boldsymbol{x}_j) = \sqrt{\boldsymbol{x}_i\,\boldsymbol{M}^{-1}\,\boldsymbol{x}_j}
\end{equation}
where $\boldsymbol{M}$ is a $D\times D$ positive definite matrix of hyperparameters.

@@

A key component of all the covariance functions discussed above is that they are all \emph{parameterized} by a set of hyperparameters.
In most cases, we would not have \emph{a priori} knowledge for how the values of these hyperparameters should be set.
Instead, their values will need to be numerically tuned or incorporated into a larger inference scheme.

@@

An important point here is that, since many data analysis procedures in astrophysics include a step like the ones listed above, the use of a GP likelihood does not significantly change the processing.
In fact, we like to say that the GP likelihood can be used as a drop-in replacement for anywhere you are currently using a ``chi-squared'' objective.
There are some practical reasons why things are not necessarily this simple (for example, computational cost, as described below), but the sentiment stands.

In this review, we will not go into too many details about the inference algorithms, but throughout the text, we will regularly use the \textsf{BFGS} gradient-based, non-linear optimization routine \citep{Nocedal:1999, scipy} to find the maximum likelihood parameter values
\begin{equation}
  \meanparams_\star,\hyperparams_\star = \argmax_{\meanparams,\hyperparams} \,\mathcal{L}(\meanparams,\hyperparams) = \argmax_{\meanparams,\hyperparams} \,\log p(\ydata\,|\,\meanparams,\hyperparams) \quad.
\end{equation}
Another common inference technique used in astrophysics---and in this review!---is the use of MCMC or nested sampling \citep{skilling2006nested} to generate posterior samples
\begin{equation}
  \meanparams,\hyperparams \sim p(\meanparams,\hyperparams\,|\,\ydata)
\end{equation}
that can be used to marginalise over some subset of the parameters, and estimate the uncertainty on the parameter values.
All the examples in this review are implemented using the \textsf{NumPyro} probabilistic inference library, and all the MCMC examples use the gradient-based No U-Turn Sampler (NUTS) algorithm \citep{Hoffman:2014}.

@@

Given the wide array of possible kernel functions described in \autoref{sec:kernels}, it can be important to assess the performance of your model and the relevant choices.
This includes both assessing the choice to use a GP in the first place, and the specific choice of kernel function.
It is important to note that there is nothing fundamentally different about GPs in this context when compared to other models for data, so that means that many methods that you may already use for model selection and validation in other contexts will also apply when using GPs.
That being said, within the astrophysics literature formal probabilistic model checking has had limited use, and GPs do come with some specific technical complications, therefore we will discuss some examples of model validation and selection techniques that have been used for GPs.

When it comes to selecting between different possible kernel functions, the approach that you take may depend on your specific research goals.
For example, in many cases, including the transiting exoplanet example in \autoref{sec:transit}, the main parameters of physical interest may be the parameters of the mean model, and the GP is simply an effective model for stochastic nuisances.
In this case, it may be sufficient to demonstrate that the inferred results are not significantly inconsistent for different choices of kernel function.

@@

GP models have been used on occasion to \textit{de-trend} light curves from transit surveys prior to running transit searches, i.e.\ to remove variability on timescales significantly longer than a transit \citep[see e.g.][]{2016ApJS..226....7C}, and are used frequently after detection to model the out-of-transit baseline alongside the transit signal itself (see \autoref{sec:transit_fit}). In principle, a GP could also be used to model out-of-transit variations (intrinsic or instrumental) alongside transits as part of a detection pipeline. Doing this simultaneously rather than sequentially allows more flexible models to be used for the out-of-transit variations, and avoids the risk of corrupting the transit signal by using an excessively aggressive filter at the detection stage, as demonstrated by \citet{2015ApJ...806..215F} for \textit{K2} data. In that case, the primary source of out-of-transit variability is instrumental systematics (see \autoref{sec:transit_fit}), which are to some degree common to all light curves in a given observing run. \citet{2015ApJ...806..215F} model these systematics as a linear combination of the first 150 principal components of the ensemble of light curves, alongside a simple box-shaped transit model, varying the latter's period, phase and duration.

In other space-based transit surveys such as \textit{CoRoT}, \emph{Kepler} and \emph{TESS}, the systematics are less prominent, so the dominant source of out-of-transit variations is intrinsic stellar variability, which occurs mainly on timescales longer than transits, and can therefore be filtered out quite effectively without removing the transits. However, in some cases, for example for rapidly rotating and magnetically active stars, the separation between transits and stellar signal becomes less clean, significantly reducing the performance of sequential de-trending approaches (irrespective of the algorithm used, \citealt{wotan}).
In such cases, a simultaneous modelling approach could improve detection performance significantly. Since the nuisance signal is specific to each star, it cannot be modelled using other light curves, but a GP is a credible alternative. Such an approach could also offer a small but significant enhancement in the sensitivity of future missions such as \textit{PLATO} to transits of habitable planets around Sun-like stars (which are very shallow and relatively long). The main issue would be computational cost: transit surveys monitor $10^4$--$10^5$ stars at a time, with $10^3$--$10^5$ observations per run, and the transit search must be run over a fine grid of periods and phases. Doing this with standard GPR methods would be impractical. However, now fast and scalable GP solvers are available, this becomes a more feasible proposition.

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but formulated in terms of wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$.
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. 
This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or \textit{TESS}.
\begin{armarginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to $\gamma=0$, $1$ and $2$, respectively.}
\end{armarginnote}
However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from ground-based, space telescopes in low-Earth orbit such as Hubble.

@@

\subsubsection{Systematics removal in the presence of stellar variability}
GP-based models have also been used to correct instrumental systematics in data from the \textit{K2} space mission, which used the \textit{Kepler} satellite to perform an Ecliptic plane survey after the failure of two of its reaction wheels. During the \textit{K2} observations, the satellite underwent significant roll-angle variations, causing the stars to move on the detector by more than a pixel over a timescale of hours. The satellite thrusters were fired every $\sim 6$\,h to return the spacecraft to its nominal attitude, but the resulting drift caused significant changes in the measured stellar fluxes, due to the detector inter- and intra-pixel variations, and to changes in the contamination of the photometric aperture by neighbouring targets as well as in aperture losses. These can be modelled effectively using a GP with a squared exponential covariance function depending on the roll angle \citep{2015MNRAS.447.2880A, 2016ApJS..226....7C} or the star's 2-D position on the detector \citep{2016MNRAS.459.2408A}. In these approaches, a second, time-dependent term is added to the GP covariance function to represent the target star's intrinsic variability. This not only improves the fit but allows the position-dependent systematics to be evaluated separately and thus removed while preserving intrinsic variability. The use of a time-dependent GP term improves the photometric performance over other widely-used methods for correcting \textit{K2} systematics \citep[e.g.,][]{2014PASP..126..948V} when the stellar variability is significant and/or occurs on timescale similar to the roll angle variations. Ultimately, the best overall photometric precision for \textit{K2} was achieved by combining PLD to model the position-dependent systematics with a time-dependent GP to model intrinsic variability \citep{everest1,everest2}.

@@

GP models were first applied to the simulated observations of the Sun's total irradiance variations by \citet{2012MNRAS.419.3147A}, to test a new method to simulate Radial Velocity (RV) variations based space-based photometry on solar data. The context of this work was the large numbers of candidate transiting exoplanets being discovered at the time by the \textit{CoRoT} and \textit{Kepler} space missions. Detecting the planet signals in RV was needed to confirm the planetary nature of the candidates by measuring their masses, but was being hampered by the apparent RV variations caused by active regions. Using simple geometric considerations, \citet{2012MNRAS.419.3147A} derived a simple (approximate) relationship between the flux perturbation caused by active regions, $F$, and their RV signature, which depends on both $F$ and its time-derivative $F'$. They tested this data-driven ``$FF'$ method'' on simulated photometric and RV observations of the Sun-as-a-star produced using resolved magnetograms, using a GP as a principled smoothing tool to evaluate $F$ and $F'$ from the photometry, and comparing the results to the simulated RVs. \citet{2012MNRAS.419.3147A}  tested a number of covariance functions, both aperiodic and Quasi-Periodic (QP), and found the former to be preferred in the solar case, as expected if most active regions on the Sun do not persist for much longer than its rotation period. More active and/or more rapidly rotating stars display variability that is coherent over multiple rotation periods, making QP GPs the model of choice. For example, \citet{2014MNRAS.443.2517H} used a QP kernel when applying the aforementioned $FF'$ method to the active planet-host star CoRoT-7.

@@

Most studies focusing on the photometric signatures of stellar activity ignore variations on short timescales, either working with binned data or using a ``jitter term'' to absorb them. This keeps the model simple and, when binning, speeds up computing time. However, when the time-sampling and precision of the observations allow it (e.g.\ for data from \textit{CHEOPS} or \textit{PLATO}), composite GP models with different terms to describe activity, granulation and stellar oscillations can be useful. For example, \citet{2020A&A...634A..75B} analysed light curves of stars observed at high cadence and high precision in the \textit{CoRoT} asteroseismology field. They showed that the parameters of planetary transits injected into these light curves were recovered more accurately when using a composite GP to model the stellar variability than one with a single term. They also tested a white noise only model, which provided less precise estimates than either GP models, albeit generally consistent with both.

\subsubsection{Stellar activity in Radial Velocities} \label{sec:RVact}
Given the remarkable instrumental precision achieved by modern RV spectrographs, stellar activity is nowadays the key factor limiting the sensitivity of RV surveys to low-amplitude and/or long-period planet signals. The effect of active regions on RV observations is two-fold. First, as dark spots rotate on the stellar surface, they distort the profile of spectral lines, removing a small contribution first from the red wing of each spectral line, then from the line core, and then from the blue wing. The second effect is more subtle: in the absence of active regions, spectral lines of Sun-like stars display a net blue-shift due to granulation: the emission from the hot, up-welling material in the granules, which is blue-shifted, dominates over that from the cooler material falling back down in the inter-granular lanes). In a facula, i.e.\ a region of enhanced magnetic flux density compared to the ``clean'' photosphere, convection is suppressed, leading to a localised reduction in this ``convective blue-shift''. Faculae have very small photometric contrast but can cover a much larger area than dark spots, so their RV signature can dominate over that of spots, especially for moderately slow rotators like the Sun with modest magnetic fields \citep{2010A&A...512A..39M}.

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/RV_framework.pdf}
  \caption{Schematic of the multidimensional GP framework of \citet{2015MNRAS.452.2269R} for modelling activity signals in RV data. The observed RVs (shown as orange points with error bars in the bottom right panel) are modelled alongside observations of activity indicators extracted from the spectra or line profiles (orange points with error bars in the bottom left and bottom middle panels). In all these time-series, the activity signal (shown by the black line in each panel) is modelled as a linear combination of an unobserved (latent) GP, shown in the top-left panel, and its first time-derivative, shown in the top middle panel. In addition, the Keplerian signal(s) of putative planet(s) (the sum of which is shown by the blue lines in the top right and bottom right panels) contribute to the RVs but not to the activity indicators. This hierarchical model is implemented in a Bayesian framework, sampling over all the hyperparameters listed in the blue boxes (subject to appropriate priors) to derive posteriors over the hyperparameters of interest and/or estimate the evidence for models with different numbers of planets. }
  \label{fig:RV_framework}
\end{figure}

@@

In certain application domains, such as the analysis of photometric and radial velocity time-series from exoplanet surveys, GPR is already used fairly widely. Its usefulness as an effective model for astrophysical or instrumental nuisance signals is well established, and it is seen as a fairly standard element of the astronomer's data analysis toolkit. In these existing applications, we foresee three main routes for further progress in the near future. One is the development of more physically motivated (as opposed to merely effective) GP models for stellar variability, so that GPR can be used to characterise and understand these stellar signals for their own sake. The other is a more systematic use of GPR as part of the de-trending and detection pipelines for exoplanet transit and RV surveys. Computational cost considerations would have made this impractical just a few years ago, but this obstacle has been lifted thanks to the recent development of scalable GP models. Finally, as JWST begins to deliver much exoplanet atmosphere observations with much higher signal-to-noise (including, importantly, more precise and relevant housekeeping parameters), a new generation of systematics models will no doubt be developed to get the most out of these observations, and it seems likely that GPR would have a role to play in that.

>>>>

In the previous section, we presented an overview of GP methods, and the key mathematical details. In this section, we will dive deeper into some of the practical decisions that arise when using GPs.
The two core elements of a GP model are the mean function $m(t;\,\meanparams)$, and the ``kernel'' or ``covariance'' function $k(t_i,\,t_j;\,\hyperparams)$. In contrast to the way GP models are typically presented in the machine learning literature, in astrophysics they will often---but not always---include non-trivial mean functions.
For example, in the case study discussed in \autoref{sec:transit}, the mean function $m(t;\,\meanparams)$ is a physical transit model that is a function of the orbital parameters of the system, and includes a realistic limb-darkening model \citep{2002ApJ...580L.171M}.
However, in this review we will not discuss the mean function in detail, focusing instead primarily on the kernel function, since that is unique to GP modelling.
All this being said, in our experience, new users of GP models will often focus and worry more than necessary about the choice of kernel function for their problem.
As with any probabilistic modelling problem, there are several well-defined workflows for motivating, selecting, and validating the choice of kernel function.
In this section we walk through this process in detail.

@@

The kernel function can be any positive scalar function that gives rise to a positive semi-definite covariance matrix over the input domain, but some will be more useful than others.
Given this large decision space, the reader may be wondering how to choose the right kernel function for their specific problem.
If one is very lucky, one may be able to motivate the choice of model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel functions have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that would naturally motivate their use in certain circumstances.
On the other hand, if the physical model is specified by its \emph{power spectrum}, this can be recast as a GP model with a specific kernel function.
% \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

Even in the absence of a formal physics-based justification for the choice of model, it may be possible to identify the key scales of the problem and design a kernel function that captures these features.
The usual approach to this problem is to take sums and products of commonly used kernel functions to construct a set of models that have the desired covariance structure, and then combine or select between these choices.
For example, if the problem of interest is expected to involve only a single non-periodic timescale, one could list all the two-parameter non-periodic kernel functions and use a numerical model selection technique as described below.

@@

When selecting a kernel function, it can be useful to generate samples from this implied prior distribution over functions to get a qualitative sense of the properties of the kernel.
In practice, this is done by choosing a grid of times $\boldsymbol{t} = \{t_i\}$, evaluating the elements of the covariance matrix
\begin{equation}
  K_{ij} = k(t_i,\,t_j;\,\hyperparams),
\end{equation}
and then generating a multivariate Gaussian sample with this covariance.
As an example, \autoref{fig:samples} shows several prior samples for three different kernel functions from \autoref{tab:kernels}, with a range of length scales $\lengthscale$ and amplitudes $\amplitude$.
In this figure, we can see some qualitative differences between the kernel functions---namely that the samples become ``smoother'' from left to right---and we can see how the characteristic input and output scales of the allowed functions change with the hyperparameters.

@@

\begin{figure}[ht]
  \centering
  \script{kernel-ops.ipynb}
  \includegraphics[width=0.7\linewidth]{figures/kernel-ops1.pdf}
  \includegraphics[width=0.7\linewidth]{figures/kernel-ops2.pdf}
  \caption{A demonstration of the effects of differentiation on a GP model.
  Each row shows the same panels for a different base kernel: \emph{(top)} a squared exponential kernel ($\lengthscale = 1$), and \emph{(bottom)} the product of a squared exponential kernel ($\lengthscale_1 = 2.5$) with a sine squared exponential kernel ($\Gamma = 1$; $\lengthscale_2 = 3.5$).
  In each row, the leftmost panel shows the kernel function and its first and second time derivatives, which are used to evaluate the covariance matrix for a GP model for a process and its derivative.
  The right-hand panel shows a realization from the process ($f(t)$, top) and its time derivative ($\dot{f}(t)$, bottom) generated by jointly sampling from the GP including the covariance between $f(t)$ and $\dot{f}(t)$.}
  \label{fig:kernel-ops}
\end{figure}

@@

When the input coordinates have some geometric interpretation (e.g., spatial coordinates) simple isotropic distance metrics, like those presented so far, are generally well-justified.
However, when the input coordinates are heterogeneous with, for example, different units, it is often more appropriate to use a more general distance metric.
In particular, it can be useful to fit for a different length scale parameter $\lengthscale_d$ for each input dimension $d$:
\begin{equation}\label{eq:ard-distance}
  \rho(\boldsymbol{x}_i,\boldsymbol{x}_j) = \sqrt{\sum_{d=1}^D \left(\frac{x_{id} - x_{jd}}{\lengthscale_d}\right)^2}
\end{equation}
or even include support for covariance between different input dimensions:
\begin{equation}
  \rho(\boldsymbol{x}_i,\boldsymbol{x}_j) = \sqrt{\boldsymbol{x}_i\,\boldsymbol{M}^{-1}\,\boldsymbol{x}_j}
\end{equation}
where $\boldsymbol{M}$ is a $D\times D$ positive definite matrix of hyperparameters.

@@

A key component of all the covariance functions discussed above is that they are all \emph{parameterized} by a set of hyperparameters.
In most cases, one would not have \emph{a priori} knowledge for how the values of these hyperparameters should be set.
Instead, their values will need to be numerically tuned or incorporated into a larger inference scheme.

@@

An important point here is that, since many data analysis procedures in astrophysics include a step like the ones listed above, the use of a GP likelihood does not significantly change the processing.
In fact, we like to say that the GP likelihood can be used as a drop-in replacement for any situation where a ``chi-squared'' objective is currently used\footnote{There may be cases where a model including correlated noise is not required to explain the data but, since the independent noise model is a special case of most GP models, using a GP isn't conceptually problematic even in those situations.} There are some practical reasons why things are not necessarily this simple (for example, computational cost, as described below), but the sentiment stands.

In this review, we will not go into too many details about the inference algorithms, but throughout the text, we will regularly use the \textsf{BFGS} gradient-based, non-linear optimization routine \citep{Nocedal:1999, scipy} to find the maximum likelihood parameter values
\begin{equation}
  \meanparams_\star,\hyperparams_\star = \argmax_{\meanparams,\hyperparams} \,\mathcal{L}(\meanparams,\hyperparams) = \argmax_{\meanparams,\hyperparams} \,\log p(\ydata\,|\,\meanparams,\hyperparams) \quad.
\end{equation}
Another common inference technique used in astrophysics---and in this review!---is the use of MCMC or nested sampling \citep{skilling2006nested} to generate posterior samples
\begin{equation}
  \meanparams,\hyperparams \sim p(\meanparams,\hyperparams\,|\,\ydata)
\end{equation}
that can be used to marginalise over some subset of the parameters, and estimate the uncertainty on the parameter values.
%All the examples in this review are implemented using the \textsf{NumPyro} probabilistic inference library, and all the MCMC examples use the gradient-based No U-Turn Sampler (NUTS) algorithm \citep{Hoffman:2014}.

@@

Given the wide array of possible kernel functions described in \autoref{sec:kernels}, it can be important to assess the performance of one's model and the relevant choices.
This includes both assessing the choice to use a GP in the first place, and the specific choice of kernel function.
It is important to note that there is nothing fundamentally different about GPs in this context when compared to other models for data. Therefore, many model selection and validation methods, which the reader may already have encountered in other contexts  \citep[see e.g.][for a useful textbook reference]{claeskens_hjort_2008}, will also apply when using GPs.
That being said, within the astrophysics literature formal probabilistic model checking has had limited use, and GPs do come with some specific technical complications, therefore we will discuss some examples of model validation and selection techniques that have been used for GPs.

When it comes to selecting between different possible kernel functions, the approach to take may depend on the specific research goals.
For example, in many cases, including the transiting exoplanet example in \autoref{sec:transit}, the main parameters of physical interest may be the parameters of the mean model, and the GP is simply an effective model for stochastic nuisances.
In this case, it may be sufficient to demonstrate that the inferred results are not significantly inconsistent for different choices of kernel function.

@@

GP models have been used on occasion to \textit{de-trend} light curves from transit surveys prior to running transit searches, i.e.\ to remove variability on timescales significantly longer than a transit \citep[see e.g.][]{2016ApJS..226....7C}, and are used frequently after detection to model the out-of-transit baseline alongside the transit signal itself (see \autoref{sec:transit_fit}). In principle, a GP could also be used to model out-of-transit variations (intrinsic or instrumental) alongside transits as part of a detection pipeline. Doing this simultaneously rather than sequentially allows more flexible models to be used for the out-of-transit variations, and avoids the risk of corrupting the transit signal by using an excessively aggressive filter at the detection stage, as demonstrated by \citet{2015ApJ...806..215F} for data from the \textit{K2}
space mission \citep{2014PASP..126..398H}, which used the \textit{Kepler} satellite to perform an Ecliptic plane survey after the failure of two of its reaction wheels. In that case, the primary source of out-of-transit variability is instrumental systematics (see \autoref{sec:transit_fit}), which are to some degree common to all light curves in a given observing run. \citet{2015ApJ...806..215F} model these systematics as a linear combination of the first 150 principal components of the ensemble of light curves, alongside a simple box-shaped transit model, varying the latter's period, phase and duration.

In other space-based transit surveys such as CoRoT citep{2006ESASP1306...33B}, \emph{Kepler} \citep{2010Sci...327..977B} and TESS \citep{2015JATIS...1a4003R}, the systematics are less prominent, so the dominant source of out-of-transit variations is intrinsic stellar variability, which occurs mainly (for main sequence stars) on timescales longer than transits, and can therefore be filtered out quite effectively without removing the transits. However, in some cases, for example for rapidly rotating and magnetically active stars, the separation between transits and stellar signal becomes less clean, significantly reducing the performance of sequential de-trending approaches (irrespective of the algorithm used, \citealt{wotan}).
In such cases, a simultaneous modelling approach could improve detection performance significantly. Since the nuisance signal is specific to each star, it cannot be modelled using other light curves, but a GP is a credible alternative. Such an approach could also offer a small but significant enhancement in the sensitivity of future missions such as PLATO \citep{2014ExA....38..249R} to transits of habitable planets around Sun-like stars (which are very shallow and relatively long). The main issue would be computational cost: transit surveys monitor $10^4$--$10^5$ stars at a time, with $10^3$--$10^5$ observations per run, and the transit search must be run over a fine grid of periods and phases. Doing this with standard GPR methods would be impractical. However, now fast and scalable GP solvers are available, this becomes a more feasible proposition.

@@

To address this, \citet{2009ApJ...704...51C} proposed a wavelet-based method to model correlated noise in transit light curves, which can be seen as a special case of GPR with a covariance function belonging to the exponential family, but formulated in terms of wavelets. Rather than modelling the covariance in the time domain, the Power Spectral Density (PSD) of the noise is assumed to be of the form $P(f) \propto f^{-\gamma}$.
Noting that the wavelet transform of such a process gives rise to a nearly diagonal covariance matrix, \citet{2009ApJ...704...51C}  derived an expression for the likelihood that can be computed in $\mathcal{O}(N)$ operations. 
This method has been used widely since its publication to model transit observations from space missions which stare continuously at a given field, such as \textit{Kepler} or TESS.
\begin{armarginnote}
  \entry{Colours of noise}{White, pink and red noise correspond to power spectral densities with power-law indices $\gamma=0$, $1$ and $2$, respectively.}
\end{armarginnote}
However, the use of a wavelet transform requires regular time sampling, which precludes its application to datasets with irregular sampling or significant data gaps, for example from ground-based, space telescopes in low-Earth orbit such as Hubble.

@@

\subsubsection{Systematics removal in the presence of stellar variability}
GP-based models have also been used to correct instrumental systematics in data from the \textit{K2} space mission \citep{2014PASP..126..398H}. During the \textit{K2} observations, the satellite underwent significant roll-angle variations, causing the stars to move on the detector by more than a pixel over a timescale of hours. The satellite thrusters were fired every $\sim 6$\,h to return the spacecraft to its nominal attitude, but the resulting drift caused significant changes in the measured stellar fluxes, due to the detector inter- and intra-pixel variations, and to changes in the contamination of the photometric aperture by neighbouring targets as well as in aperture losses. These can be modelled effectively using a GP with a squared exponential covariance function depending on the roll angle \citep{2015MNRAS.447.2880A, 2016ApJS..226....7C} or the star's 2-D position on the detector \citep{2016MNRAS.459.2408A}. In these approaches, a second, time-dependent term is added to the GP covariance function to represent the target star's intrinsic variability. This not only improves the fit but allows the position-dependent systematics to be evaluated separately and thus removed while preserving intrinsic variability. The use of a time-dependent GP term improves the photometric performance over other widely-used methods for correcting \textit{K2} systematics \citep[e.g.,][]{2014PASP..126..948V} when the stellar variability is significant and/or occurs on timescale similar to the roll angle variations. Ultimately, the best overall photometric precision for \textit{K2} was achieved by combining PLD to model the position-dependent systematics with a time-dependent GP to model intrinsic variability \citep{everest1,everest2}.

@@

GP models were first applied to the simulated observations of the Sun's total irradiance variations by \citet{2012MNRAS.419.3147A}, to test a new method to simulate Radial Velocity (RV) variations based space-based photometry on solar data. The context of this work was the large numbers of candidate transiting exoplanets being discovered at the time by the CoRoT and \textit{Kepler} space missions. Detecting the planet signals in RV was needed to confirm the planetary nature of the candidates by measuring their masses, but was being hampered by the apparent RV variations caused by active regions. Using simple geometric considerations, \citet{2012MNRAS.419.3147A} derived a simple (approximate) relationship between the flux perturbation caused by active regions, $F$, and their RV signature, which depends on both $F$ and its time-derivative $F'$. They tested this data-driven ``$FF'$ method'' on simulated photometric and RV observations of the Sun-as-a-star produced using resolved magnetograms, using a GP as a principled smoothing tool to evaluate $F$ and $F'$ from the photometry, and comparing the results to the simulated RVs. \citet{2012MNRAS.419.3147A}  tested a number of covariance functions, both aperiodic and Quasi-Periodic (QP), and found the former to be preferred in the solar case, as expected if most active regions on the Sun do not persist for much longer than its rotation period. More active and/or more rapidly rotating stars display variability that is coherent over multiple rotation periods, making QP GPs the model of choice. For example, \citet{2014MNRAS.443.2517H} used a QP kernel when applying the aforementioned $FF'$ method to the active planet-host star CoRoT-7.

@@

Most studies focusing on the photometric signatures of stellar activity ignore variations on short timescales, either working with binned data or using a ``jitter term'' to absorb them. This keeps the model simple and, when binning, speeds up computing time. However, when the time-sampling and precision of the observations allow it (e.g.\ for data from \textit{CHEOPS}, \citealt{2021ExA....51..109B}, or PLATO), composite GP models with different terms to describe activity, granulation and stellar oscillations can be useful. For example, \citet{2020A&A...634A..75B} analysed light curves of stars observed at high cadence and high precision in the CoRoT asteroseismology field. They showed that the parameters of planetary transits injected into these light curves were recovered more accurately when using a composite GP to model the stellar variability than one with a single term. They also tested a white noise only model, which provided less precise estimates than either GP models, albeit generally consistent with both.

\subsubsection{Stellar activity in Radial Velocities} \label{sec:RVact}
Given the remarkable instrumental precision achieved by modern RV spectrographs, stellar activity is nowadays the key factor limiting the sensitivity of RV surveys to low-amplitude and/or long-period planet signals. The effect of active regions on RV observations is two-fold. First, as dark spots rotate on the stellar surface, they distort the profile of spectral lines, removing a small contribution first from the blue wing of each spectral line, then from the line core, and then from the red wing. The second effect is more subtle: in the absence of active regions, spectral lines of Sun-like stars display a net blue-shift due to granulation: the emission from the hot, up-welling material in the granules, which is blue-shifted, dominates over that from the cooler material falling back down in the inter-granular lanes. In a facula, i.e.\ a region of enhanced magnetic flux density compared to the ``clean'' photosphere, convection is suppressed, leading to a localised reduction in this ``convective blue-shift''. Faculae have very small photometric contrast but can cover a much larger area than dark spots, so their RV signature can dominate over that of spots, especially for moderately slow rotators like the Sun with modest magnetic fields \citep{2010A&A...512A..39M}.

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/RV_framework.pdf}
  \caption{Schematic of the multidimensional GP framework of \citet{2015MNRAS.452.2269R} for modelling activity signals in RV data. The observed RVs (shown as orange points with error bars in the bottom right panel) are modelled alongside observations of activity indicators extracted from the spectra or line profiles (orange points with error bars in the bottom left and bottom middle panels). In all these time-series, the activity signal (shown by the black line in each panel) is modelled as a linear combination of an unobserved (latent) GP, shown in the top-left panel, and its first time-derivative, shown in the top middle panel. In addition, the Keplerian signal(s) of putative planet(s) (the sum of which is shown by the blue lines in the top right and bottom right panels) contribute to the RVs but not to the activity indicators. This hierarchical model is implemented in a Bayesian framework, sampling over all the hyperparameters listed in the blue boxes (subject to appropriate priors) to derive posteriors over the hyperparameters of interest and/or estimate the evidence for models with different numbers of planets. Models with different numbers of planets (or without planets) are then compared using the methods discussed in Section~\ref{sec:model_selection}.}
  \label{fig:RV_framework}
\end{figure}

@@

In certain application domains, such as the analysis of photometric and radial velocity time-series from exoplanet surveys, GPR is already used fairly widely. Its usefulness as an effective model for astrophysical or instrumental nuisance signals is well established, and it is seen as a fairly standard element of the astronomer's data analysis toolkit. In these existing applications, we foresee three main routes for further progress in the near future. One is the development of more physically motivated (as opposed to merely effective) GP models for stellar variability, so that GPR can be used to characterise and understand these stellar signals for their own sake. The other is a more systematic use of GPR as part of the de-trending and detection pipelines for exoplanet transit and RV surveys. Computational cost considerations would have made this impractical just a few years ago, but this obstacle has been lifted thanks to the recent development of scalable GP models. Finally, as JWST \citep{2006SSRv..123..485G} begins to deliver much exoplanet atmosphere observations with much higher signal-to-noise (including, importantly, more precise and relevant housekeeping parameters), a new generation of systematics models will no doubt be developed to get the most out of these observations, and it seems likely that GPR would have a role to play in that.
